---
title: "Day 1 Exercises Kristian Nikolov"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the data

###Loading the gapminder data using read_csv

```{r Loading gapminder}
setwd("C:/Northwestern MSiA/Bootcamp/R Training/R-bootcamp-2019")
gapminder <- read.csv("data/gapminder5.csv", stringsAsFactors = FALSE)
```

## If Statement

Using an if() statement to print a suitable message reporting whether there are any records from 2002 in the gapminder dataset. 

```{r If statement for records from 2002}
if (any(gapminder$year==2002)){
  print("There are records from 2002")
} else {
  print("Sorry, no records exist!")
}
```

Now doing the same for 2012.
```{r If statement for records from 2012}
if (any(gapminder$year==2012)){
  print("There are records from 2012")
} else {
  print("Sorry, no records exist!")
}
```

## Loop and If Statements

###Writing a script that finds the mean life expectancy by country for countries whose population is below the mean for the dataset

```{r MLE for countries where population<mean}
avg_pop<-mean(gapminder$pop)

for (i in unique(gapminder$country)){
  tmp<- mean(gapminder$pop[gapminder$country==i])
  mle<- mean(gapminder$lifeExp[gapminder$country==i])
  
  if (tmp<avg_pop) {
    print(paste0("MLE in ", i, " is ", mle))
  }
}
```

###Writing a script that loops through the gapminder data by continent and prints out whether the mean life expectancy is smaller than 50, between 50 and 70, or greater than 70.

```{r Checking whether continent MLE is <50, b/n 50 and 70, or >70 }
for (i in unique(gapminder$continent)){
  mle<- mean(gapminder$lifeExp[gapminder$continent==i])
  
  if (mle<50) {
    print(paste("MLE in", i, "is smaller than 50"))
  } else {
    if (mle>70) {
      print(paste("MLE in", i, "is greater than 70"))
    } else {
      print(paste("MLE in", i,"is between 50 and 70"))
    }
  }
}
```

##Exercise: Write Functions

###Creating a function that given a data frame will print the name of each column and the class of data it contains. Testing with the gapminder dataset.

```{r Function which gives a data frames col_names and their data classes}
col_names_and_classes<-
  function(df){
    for(i in names(df)){
      d_class<-class(df[,i])
      print(paste("Column name:", i, "; Data class:", d_class))
    }
  }

col_names_and_classes(gapminder)
```

###Creating a function that given a vector will print the mean and the standard deviation of a vector, it will optionally also print the median.

```{r Function which given a vector prints its mean, sd and median}
vector_stats<-
  function(vec, vec_median=FALSE){
    if (vec_median) {
      cat("Mean:", mean(vec),
          "\nStandard Deviation:", sd(vec),
          "\nMedian:", median(vec))
    } else {
      cat("Mean:", mean(vec),
          "\nStandard Deviation:", sd(vec), "\n")
    }
  }
vector_stats(gapminder$pop)
vector_stats(gapminder$pop, vec_median = TRUE)
```

##Analyzing the relationship

###What is the relationship between GDP per capita and life expectancy? Does this relationship change over time?

Since GDP/capita is a much larger number than life expectancy, we can use the natural log of both variables to observe if their is a relationship in the ways they change.

```{r Relationship between GDP per capita and life expectancy}
log_GDP<-log(gapminder$gdpPercap)
log_le<-log(gapminder$lifeExp)

cor1 <- round(cor(log_GDP, log_le),digits=2) #0.78 correlation is extremely high. Let's plot them against each other to observe it better
plot(log_GDP ~ log_le,
     main = paste("Scatterplot of GDP per capita and Life Expectancy, \nCorrelation =", cor1),
     xlab = "ln of Life Expectancy", ylab = "ln of GDP per capita")
```

As can be seen from the plot, the higher GDP per capita is, the more life expectancy increases.

###Does the relationship between GDP per capita and life expectacy vary by continent? 
```{r Relationship between GDP per capita and life expectancy by continent}
#First we need to split North and South America
gapminder_recoded <- gapminder
gapminder_recoded$continent[gapminder_recoded$continent=="Americas"] <- "South America"
gapminder_recoded$continent[gapminder_recoded$country=="Canada"] <-  "North America"
gapminder_recoded$continent[gapminder_recoded$country=="Mexico"] <-  "North America"
gapminder_recoded$continent[gapminder_recoded$country=="United States"] <-  "North America"

#Now we need to check previous exercise for each continent
for (i in unique(gapminder_recoded$continent)){
log_GDP2<-log(gapminder_recoded$gdpPercap[gapminder_recoded$continent==i])
log_le2<-log(gapminder_recoded$lifeExp[gapminder_recoded$continent==i])

cor2 <- round(cor(log_GDP2, log_le2), digits = 2)
sd_gdp <- round(sd(gapminder_recoded$gdpPercap[gapminder_recoded$continent==i]))

#print(paste("In", i, "the correlation between GDP and LE is", cor2, "while the GDP standard deviation is", #sd_gdp)) - this can be used, but I found it redundant after inserting these in the plot titles

plot(log_GDP2 ~ log_le2,
     main = paste("Scatterplot of GDP per capita and Life Expectancy in", i,", \nCorrelation =", cor2, "GDP standard deviation is",sd_gdp),
     xlab = "ln of Life Expectancy", ylab = "ln of GDP per capita")
}
```
The correlation between GDP and life expectancy (LE) is strongest in Oceania and America (where we can also observe a very large wage gap), while it is weakest in Africa (where the wage gap is the smallest). In Asia the wage gap is the highest, whilst GDP and LE are not that correlated (many rich countries have lower GDP), signalling that there is probably something else affecting LE